% vim:ft=tex
\section{Related Work}

\subsection{Types and Programming Languages}

Types and Programming Languages~\cite{pierce2002types} provides a comprehensive
introduction to type system and programming
language theory. It introduces the basic concepts and notational conventions
required to comprehend anything slightly more advanced related to type systems.
It provided an intuitive and formal description of System F and of
Hindley-Milner, both of which are fundamental for understanding Haskell's type
system and it's compilation process.

\subsection{System F with Type Equality Coercions}

System F with Type Equality Coercions\cite{Sulzmann:2007:SFT:1190315.1190324}
originally introduced System $F_C$. Even though the version of System $F_C$ used
in the implementation deviates slightly from the one presented in this paper, it
still serves as a more complete reference and as a nice introduction to the key
idea's of System $F_C$.

\subsection{How to make \textit{ad-hoc} polymorphism less \textit{ad hoc}}

How to make \textit{ad-hoc} polymorphism less
\textit{ad-hoc}\cite{Wadler:1989:MAP:75277.75283} is the original
proposal for ad-hoc polymorphism by the use of type classes and mainly serves a
historical purpose. It provides some intuition for reasoning about type classes
and their translation into System F as dictionaries.

\subsection{Type Classes Functional with Dependencies}

Type Classes with Functional Dependencies\cite{Jones00typeclasses} nicely
illustrates the problems with multi-parameter type classes and ambiguities, how
functional dependencies can solve these issues, and how the proposed solution
was derived from relational database theory. It provides a very high level
sketch on how one might implement a type inference algorithm that supports
functional dependencies.

\subsection{Elaboration on Functional Dependencies}

Elaboration on Functional
Dependencies\cite{Karachalias:2017:EFD:3156695.3122966} paper contains the
specification of the type inference algorithm and elaboration into System $F_C$
that will be used for the implementation that will be presented by this thesis.
This is the most important related work as most of the work presented in this
thesis will be based upon this.

Aditionally, this paper provides a specification for a variant of System $F_C$
that is currently used to implement System $F_C$ in the implementation.

\subsection{Quantified Class Constraints}

The relation to Quantified Class
Constraints\cite{Bottu:2017:QCC:3156695.3122967} is artificial. The original
prototype compiler my work is based on was built to support the type system
feature that is presented in this paper. This paper served as a specification
that facilitated easier navigation through the existing codebase as the code
closely corresponded to the presented theory.

\subsection{OutsideIn(X)}

OutsideIn(X)\cite{outsideinx-modular-type-inference-with-local-assumptions}
served as an excellent guide to the current state of affairs on how type
inference is done in GHC. And also as an introduction to the general structure
of more advanced type inference algorithms. Lastly, this paper describes a type
inference algorithm that supports both GADTs and Type Families and will
therefore be used as a specification to implement both of these features in the
prototype compiler to study the interaction between these and our
implementation.
